Sparse Function-space Representation \\ of Neural Networks

### 2023-05-15 paper/main.tex
Sec. 3
updated rl experiemetn

\paragraph{Sparsifying the NNGP}
\label{sec:sparse-dual-gp}
%
Sparse Gaussian processes reduce the computational complexity by representing the GP as a low-rank approximation induced by a sparse set of inputs points (see \cite{quinonero2005unifying} for an early overview). Given that we have computed the dual variables derived from our neural network predictions and a kernel function, we could essentially employ any of these sparsification methods. In this work, we opt for the approach suggested by \citet{titsias2009variational} (also used in the DTC approximation, see \cite{quinonero2005unifying}), which defines the marginal predictive distribution as $q_{\vu}(f_i)  = \int p(f_i  \mid \vu) \, q(\vu) \, \mathrm{d}\vu$. Given that we have $p(f_i \mid \vu)$ determined by our GP prior, the goal is to find a $q(\vu)$.

As demonstrated in \cite{adam2021dual}, the posterior under this model bears a structure akin to \cref{eq:gp_pred}. The authors of that paper see the significance of the dual variables for approximate inference, but write them using the natural parameterization, primarily because this form is more suitable for optimization through natural gradients. In order to link the dual variables defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual variables
%
\begin{equation} 
  \valpha_{\vu}  =  \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \hat{\alpha}_{i}
  \quad \text{and} \quad
  \vbeta_{\vu} =  \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
\label{eq:dual_sparse}
\end{equation}
%
where the sparse dual variables are now a sum over all data points, with $\valpha_{\vu} \in \R^{M}$ and $\vbeta_{\vu} \in \R^{M  \times M}$. Using these sparse definition of the dual variables, our sparse GP posterior takes the following form:
\begin{equation}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] = \vkzs^{\T} \MKzz^{-1}   \valpha_{\vu} , \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta_{\vu})^{-1} ]\vkzs
\end{equation}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\MZ \in \R^{M \times D} $  instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged neural network are $(\hat{\alpha}_i, \hat{\beta}_i)$ \cref{eq:dual_param_laplac}\todo{these variables need another subscript for differentiate that from the values computed at the data points} and a kernel function $\kappa$ . Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(n^3)$ to $\mathcal{O}(m^3)$, with $m \ll n$.  Crucially, given the structure of our probabilistic model, our sparse dual variables \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

@@

A key challenge in RL is balancing the trade-off between exploration and exploitation.
That is, should an agent select actions that it knows will lead to high reward (exploitation), or should it
select new ones in hope to discover actions leading to higher reward (exploration).
One promising direction is to model the uncertainty associated with a learned transition dynamics and use it to guide exploration.
Prior work has learned dynamics models using GPs \cite{deisenrothPILCO2011,kamtheDataEfficient2018},
ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
and variational inference \cite{galImproving2016,houthooftVIME2017}.
There are then many ways to leverage uncertainty in model-based RL, for example,
taking an expectation over epistemic uncertainty  \cite{deisenrothPILCO2011,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
sampling from the posterior, akin to Thompson sampling but referred to as posterior sampling RL
\cite{osbandMoreEfficientReinforcement2013},
and methods based on upper confidence bounds (UCB) \cite{curiEfficient2020}.

% Uncertainty-guided decision making strategies
We refer the reader to \cref{app:rl} for an overview of the reinforcement learning problem and more details of our experiments.

We test \our by using it to learn a dynamics model
our method on the cart pole swing up task in MuJoCo \cite{todorov2012mujoco}, a classic benchmark for nonlinear control, see \cref{fig:rl}.
The goal is to swing the pole up and balance it around the upward position.
We increase the  difficulty of exploration in this environment by using a sparse reward function.

We compare \our (cyan) to a Laplace-GGN with GLM predictions (red), an ensemble of neural networks (green) and a basic MLP wih no uncertainty (magenta).
To ensure a fair comparison, we use the same MLP architecture/training scheme and use them in the same model-based RL algorithm, detailed in \cref{app:rl}.
% See \cref{app:rl} for more details of our experiments.
We also compare to deep deterministic policy gradient (DDPG) \cite{lillicrapContinuousControlDeep2016}, a model-free RL algorithim (yellow).
% As expected DDPG is the least sample inefficient
The training curves in \cref{fig:rl} show that \our's uncertainty estimates are useful for guiding exploration as it converges in fewer episodes.

@@

\begin{figure}[!t]
 \centering
 \begin{subfigure}[c]{.24\textwidth}
 \resizebox{\textwidth}{!}{%
 \begin{tikzpicture}[inner sep=0,outer sep=0]

>>>>

\paragraph{Sparsifying the NNGP}
\label{sec:sparse-dual-gp}
%
Sparse Gaussian processes are a way to reduce the computational complexity of a full GP using a low-rank approximation. Many sparsifying approaches exist and we refer the reader to \cite{quinonero2005unifying} for an overview. Given that we computed the dual variables derived from our neural network predictions and a kernel function, we could essentially use any approximation.\todo{approximation to get what? kernel and dual variables} We decide to follow the sparsifying approach of \citet{titsias2009variational} and also used in the DTC approximation \citep{quinonero2005unifying}, which defines the marginal predictive distribution as $q_{\vu}(f_i)  = \int p(f_i  \mid \vu) q(\vu)  d\vu$. Given that we have $p(f_i  \mid \vu)$ determined by our GP prior, the goal is to find a $q(\vu)$.

As was shown in \cite{adam2021dual}, the posterior under this model has a particular structure similar to \cref{eq:gp_pred}. In the paper, the authors spot the dual variables' importance for approximate inference, but write them with the natural parameterization as this form is more suitable for optimization through natural gradients. In order to link the dual variables defined in \cref{eq:dual_param}, we write them in the sparse GP using the dual variables
\begin{equation} 
	\valpha_{\vu}  =  \sum_{i \in \mathcal{D}_{n}}  \vkzi \, \hat{\alpha}_{i}, \quad
	\vbeta_{\vu} =  \sum_{i \in \mathcal{D}_{n}} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
\label{eq:dual_sparse}
\end{equation}
where the sparse dual variables are now a sum over all data points, with $\valpha_{\vu} \in \R^{M}$ and $\vbeta_{\vu} \in \R^{M  \times M}$. Using these sparse definition of the dual variables, our sparse GP posterior takes the following form:
\begin{equation}\label{eq:dual_sparse_post}
   \myexpect_{q_{\vu}(\vf)}[f_i] = \vkzs^{\T} \MKzz^{-1}   \valpha_{\vu} , \quad 
   \textrm{Var}_{q_{\vu}(\vf)}[f_i]  = \kappa_{ii} - \vkzs^\top [\MKzz^{-1} - (\MKzz + \vbeta_{\vu})^{-1} ]\vkzs
\end{equation}
where $\MKzz$ and $\vkzs$ are defined similarly to $\MKxx$ and $\vk_{\vx i}$ but over the inducing points $\MZ \in \R^{M \times D} $  instead of the full data set $\dataset$. The key quantities we need to make predictions from our sparse GP from the converged neural network are $(\hat{\alpha}_i, \hat{\beta}_i)$ \cref{eq:dual_param_laplac}\todo{these variables need another subscript for differentiate that from the values computed at the data points} and a kernel function $\kappa$ . Contrasting \cref{eq:dual_sparse_post} and \cref{eq:gp_pred}, we can see that the computational complexity went from $\mathcal{O}(n^3)$ to $\mathcal{O}(m^3)$, with $m \ll n$.  Crucially, given the structure of our probabilistic model, our sparse dual variables \cref{eq:dual_sparse} are a compact representation of the full model projected using the kernel.

@@

% A key challenge in RL is balancing the trade-off between exploration and exploitation.
% That is, should an agent select actions that it knows will lead to high reward (exploitation), or should it
% select new ones in hope to discover actions leading to higher reward (exploration).
% One promising direction is to model the uncertainty associated with a learned transition dynamics and use it to guide exploration.
% Prior work has learned dynamics models using GPs \cite{deisenrothPILCO2011,kamtheDataEfficient2018},
% ensembles of neural networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
% and variational inference \cite{galImproving2016,houthooftVIME2017}.
% There are then many ways to leverage uncertainty in model-based RL, for example,
% taking an expectation over epistemic uncertainty  \cite{deisenrothPILCO2011,kamtheDataEfficient2018,chuaDeepReinforcementLearning2018},
% sampling from the posterior, akin to Thompson sampling but referred to as posterior sampling RL
% \cite{osbandMoreEfficientReinforcement2013},
% and methods based on upper confidence bounds (UCB) \cite{curiEfficient2020}.

In this section, we show that \our's uncertainty estimates can be used to guide exploration in model-based RL;
demonstrating the quality of our uncertainty estimates.
% Uncertainty-guided decision making strategies
We test \our by using it to learn a dynamics model within a model-based RL strategy which uses posterior sampling to guide exploration
\cite{osbandMoreEfficientReinforcement2013,osbandWhyPosteriorSampling2017}.
We use the cart pole swing up task in MuJoCo \cite{todorov2012mujoco}, a classic benchmark for nonlinear control, see \cref{fig:rl}.
The goal is to swing the pole up and balance it around the upward position.
We increase the difficulty of exploration by using a sparse reward function.
We refer the reader to \cref{app:rl} for an overview of the reinforcement learning problem, details of the algorithim we implement and our experiment's configuration.

\cref{fig:rl} shows training curves using \our as the dynamic model (cyan) as well as a Laplace-GGN with GLM predictions (red), an ensemble of neural
networks (green) and a basic MLP wih no uncertainty (magenta).
To ensure a fair comparison, we use the same MLP architecture/training scheme and use them in the same model-based RL algorithm, detailed in \cref{app:rl}.
% that \our's uncertainty estimates are useful for guiding exploration as it converges in fewer episodes.
We also compare to deep deterministic policy gradient (DDPG) \cite{lillicrapContinuousControlDeep2016}, a model-free RL algorithim (yellow).
The training curves show that \our's uncertainty estimates are useful for guiding exploration as it converges in fewer episodes, i.e.
it is more sample efficient.
As expected, the MLP strategy (without uncertainty) was not able to successfuly explore the environment.

% We compare \our (cyan) to a Laplace-GGN with GLM predictions (red), an ensemble of neural networks (green) and a basic MLP wih no uncertainty (magenta).
% To ensure a fair comparison, we use the same MLP architecture/training scheme and use them in the same model-based RL algorithm, detailed in \cref{app:rl}.
% See \cref{app:rl} for more details of our experiments.
% As expected DDPG is the least sample inefficient
% The training curves in \cref{fig:rl} show that \our's uncertainty estimates are useful for guiding exploration as it converges in fewer episodes.

@@

 \todo{I'm thinking we should take out the SFR with updates result}
\begin{figure}[!t]
 \centering
 \begin{subfigure}[c]{.24\textwidth}
 \resizebox{\textwidth}{!}{%
 \begin{tikzpicture}[inner sep=0,outer sep=0]
